---
title: "popbio_demo_short"
format: html
editor: visual
---

# 1. Load packages

```{r}
install.packages("popbio")

library(tidyverse)
library(popbio)
```

# 2. Load census data

Use pre-loaded test.census data, which reports 3 years of census observations for a hypothetical plant.

```{r}
data <- test.census
```

# 3. Construct Projection Matrix

## A. Create a stage-fate dataframe.

A stage-fate dataframe links each observation of an individual during one census (time 1) with its fate in the following census (time 2).

```{r}
# create the stage-fate dataframe
trans <- subset(
  merge(data, data, by = "plant", sort = FALSE),
  year.x == year.y - 1)

# rename rows to improve clarity
rownames(trans) <- 1:nrow(trans)

# rename columns 2-7 to improve clarity
colnames(trans)[2:7] <- c("year", "stage", "fruits", "year2", "fate", "fruits2")
```

## B. Calculate and add fertility rate to the stage-fate data frame.

Calculating fertility rates will allow us to determine the reproductive success of each individual. It is also necessary to produce the projection matrix when individual fertility rates are averaged by stage. When census data lacks information on the direct number of offspring (ie. seeds produced by plants), a relative fertility rate could be used instead. The relative fertility is calculcated by finding the proportion of total reproductive output attributable to an individual times the total number of seedlings at the end of the projection interval.

```{r}
#########################
##### FOR YEAR 2001 #####
#########################

# subset the data from 2001
trans01 <- subset(trans, year == 2001, c(plant, stage, fruits, fate))

# count the number of seedlings observed in the next year (2002)
seedlings02 <- nrow(subset(trans, year == 2002 & stage == "seedling")) 

# calculate the relative fertility rate of each plant and store it in a new "seedling" column
trans01$seedling <- trans01$fruits/sum(trans01$fruits) * seedlings02

```

```{r}
#########################
##### FOR YEAR 2002 #####
#########################

# subset the data from 2002
trans02 <- subset(trans, year == 2002, c(plant, stage, fruits, fate))

# count the number of seedlings observed in the next year (2003)
seedlings03 <- nrow(subset(trans, year == 2003 & stage == "seedling")) 

# calculate the relative fertility rate of each plant and store it in a new "seedling" column
trans02$seedling <- trans02$fruits/sum(trans02$fruits) * seedlings03
```

## C. Create the projection matrix

```{r}
#########################
##### FOR YEAR 2001 #####
#########################

# assign life stages of interest to "stages"
stages <- c("seedling", "vegetative", "reproductive")

# create population projection matrix using 2001 relative fertility rate
A01 <- projection.matrix(trans01, sort = stages)
```

```{r}
#########################
##### FOR YEAR 2002 #####
#########################

# assign life stages of interest to "stages"
stages <- c("seedling", "vegetative", "reproductive")

# create population projection matrix using 2001 relative fertility rate
A02 <- projection.matrix(trans02, sort = stages)

```

# 4. Calculate Deterministic Population Growth Rate

The deterministic approach to population modeling relies on the equation Nt+1. It assumes density-independent population growth, constant vital rates, constant environment.

```{r}
#####################################
##### USING 2001 FERTILITY RATE #####
#####################################

# define starting population size AKA initial stage vector. the initial stage vector is th distribution of individuals across different stages at the beginning of the projection period.

# count the number of seedlings, vegetative, and reproductive individuals observed in 2001
n2001 <- data.frame(
  seedling = nrow(subset(trans, year == 2001 & stage == "seedling")),
  vegetative = nrow(subset(trans, year == 2001 & stage == "vegetative")),
  reproductive = nrow(subset(trans, year == 2001 & stage == "reproductive"))
)

# assign the starting population size. 6 seedlings, 2 vegetative, 3 reproductive
n01 <- c(6,2,3)

# project population growth over 15 years
p01 <- pop.projection(A01, n01, 15)

```

```{r}
#####################################
##### USING 2002 FERTILITY RATE #####
#####################################

# assign the starting population size. 6 seedlings, 2 vegetative, 3 reproductive
n02 <- c(6,2,3)

# project population growth over 15 years
p02 <- pop.projection(A02, n02, 15)
```

# 5. Calculate Stochastic Growth Rate

```{r}
# find stochastic growth rate for 50,000 iterations

#compile all projection matrices into one list
matrices <- list(A01, A02)
names(matrices) <- c("A01", "A02")

# find stochastic growth rate
sgr <- stoch.growth.rate(matrices, verbose = FALSE)
exp(sgr$approx)

sgr
```

### Homework Assignment

Using the prepared hudsonia matrices available in popbio, answer the following: ● What is the population growth rate for each year of available data (1985-1988)? Use the deterministic approach. ● What is the stochastic growth rate? ● Compare the growth rates calculated using the two methods. How are they different? Why?

```{r}

library(tidyverse)
library(popbio)
```

```{r}

#Load in data
hudsonia_data <- hudsonia
view(hudsonia_data)

#create a dataframe from the matrix for each year
hudsonia_85 <- hudsonia$A85
hudsonia_86 <- hudsonia$A86
hudsonia_87 <- hudsonia$A87
hudsonia_88 <- hudsonia$A88

#calculate growth rate for each year

growth_rate_85 <- lambda(hudsonia_85)
growth_rate_85

growth_rate_86 <- lambda(hudsonia_86)
growth_rate_86

growth_rate_87 <- lambda(hudsonia_87)
growth_rate_87

growth_rate_88 <- lambda(hudsonia_88)
growth_rate_88
```

● What is the population growth rate for each year of available data (1985-1988)? Use the deterministic approach. Growth Rate 1985 - 0.9593438 Growth Rate 1986 - 1.009809 Growth Rate 1987 - 0.8453119 Growth Rate 1988 - 1.01832

```{r}
##Stochastic growth rate

sgr <- stoch.growth.rate(hudsonia_data, verbose = FALSE)
exp(sgr$approx)




```

● What is the stochastic growth rate?

0.9635547

● Compare the growth rates calculated using the two methods. How are they different? Why?

They are pretty similar, the deterministic model varies by year, sometimes above one, sometimes below 1. Stochastic number is likely more realistic as it takes into account fluctuations year to year, and also the randomness of population dynamics. The year to year predictions indicate some population growth over the years, while the stochastic model predicts slow population decline at a rate close to a sustaining rate.
